---
title: "WBC_analysis"
author: "Johannes Gawron"
date: "2025-01-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Role of WBCs in oligoclonality and genetic diversity
Here, we study the role of white blood cells in the formation for the genetic
diversity respectively the oligoclonality of CTC clusters.

# Loading data

```{r loading, include=FALSE}
data <- read_tsv("~/work/ctc-data/WES_experiment/splitting_summaries/splittingSummary_full_with_sample_names_annotated.tsv")

data <- data %>% mutate(WBC = n_wbcs > 0)

data <- data %>% mutate(impact_mutations = high_impact_mutations + medium_impact_mutations)
```
Here is an overview of the data:

```{r data overview, include=TRUE}
print(data)
```
# Preprocessing

The data from Br26 will be excluded manually, as this is in fact not a breast
cancer sample but an ovarian cancer sample (which we don't study here).

```{r filtering, include=TRUE}
data <- data %>% dplyr::filter(`Sample Name` != "Br26")
filtered_data <- data %>%
  filter(str_detect(`Sample Name`, "Br|Pr|LM2"))
```

# Data analysis
To ensure that our analysis is not confounded, we control for the number of
cells in the CTC clusters and the sample identity throughout the analyses.

## Predicting oligoclonality from the presence of white blood cells

```{r fit1, include=TRUE}
fit <-
  glm(
    as.factor(Oligoclonal) ~ n_cells + `Sample Name` + WBC,
    data = filtered_data,
    family = binomial(link = "logit")
  )
summary(fit)
```
There is no significant effect of the presence of white blood cells on the
oligoclonality status of the CTC clusters was found.

## Effect of the number of white blood cells on Oligoclonality

```{r fit2, include=TRUE}
fit <-
  glm(
    as.factor(Oligoclonal) ~ n_cells + `Sample Name` + n_wbcs,
    data = filtered_data,
    family = binomial(link = "logit")
  )
summary(fit)
```
There is no significant effect of the number of white blood cells on the
oligoclonality status of the CTC clusters wasfound.


## Predicting the presence of mutations with high functional impact
```{r fit}
fit <-
  glm(
    as.factor(Oligoclonal) ~ high_impact_mutations + n_cells + `Sample Name`,
    data = filtered_data,
    family = binomial(link = "logit")
  )
summary(fit)
```
No significant effect was found for predicting the number of mutations with a
high functional impact as a function of the presence of white blood cells and of
the clonality status.


## Predicting oligoclonality from the number of mutations with functional impact

```{r fit4}
fit <-
  glm(
    as.factor(Oligoclonal) ~ n_cells + `Sample Name` + high_impact_mutations,
    data = filtered_data,
    family = binomial(link = "logit")
  )
summary(fit)
```

## Predicting oligoclonality from the number of driver mutations
```{r fit5}
fit <-
  glm(
    as.factor(Oligoclonal) ~ n_cells + `Sample Name` + drivers,
    data = filtered_data,
    family = binomial(link = "logit")
  )
summary(fit)
```
