---
title: "__tree__"
author: "__author__"
date: "__date__"
output:
  html_document:
    keep_md: yes
params:
  inputFolder: "/Users/jgawron/Documents/projects/CTC_backup/input_folder"
  nSamplingEvents: 1000
  simulationInputFolder: "/Users/jgawron/Documents/projects/
  CTC_backup/simulations/simulations2"
  treeName: "Br16_B"
  functionsScript: "/Users/jgawron/Documents/projects/CTC-SCITE/
  CTC-SCITE/experiments/workflow/resources/functions.R"
  
  
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Splitting statistics

This code analyses splitting statistics for CTC-clusters.

The analysis takes a list of trees sampled from its posterior distribution as
input and computes the mutations placement probability distribution for each one
of them. From this distribution we derive a score that quantifies the
probability that two cells have experienced divergent evolution. This score is
called the splitting score.


## Configure the script
```{r config}
inputFolder <- params$inputFolder
simulationInputFolder <- params$simulationInputFolder
treeName <- params$treeName
nTreeSamplingEvents <- params$nSamplingEvents
nMutationSamplingEvents <- params$nSamplingEvents
nSamplingEvents <- params$nSamplingEvents
```

## Loading data
```{r load, results="hide"}
source(params$functionsScript)


input <- load_data(inputFolder, treeName)
```


## Sample description 

Each row corresponds to a cell.
Column description:
  - Cluster: An number indicating the sample the cell belongs to.
  - ClusterName: The name of the sample in the nodeDescription.tsv file
  - WBC: a binary vector indicating whether the cell is a white blood cell (1)
  or not (0).
  - color: Indicates the color of the cluster in the tree, as described in the
  nodeDescription.tsv
          file.
  
```{r sample-description}
print(input$sample_description)
```



## General overview

```{r overview}
print(sprintf("We sample %d trees", nSamplingEvents))
```

For each pair of cells in the same cluster and each sampled tree we compute the
splitting score, that is, the probability that the two cells have experienced
divergent evolution. A low splitting score (close to 0) indicates that the two
cells are likely genealogically closely related, while a high splitting score
(close to 1) indicates that the two cells have evolved in a divergent manner.

Throughout the sampling of trees, this gives rise to an empirical distribution
of splitting scores for each pair of cells in the same cluster. Intuitively,
this distribution takes into account the uncertainty in the tree estimation.
To be able to interpret the splitting score appropriately (e.g. to answer the
question when is a splitting score is high enough to call oligo-clonality) we
need to calibrate our expectations.

We do this by assessing the distributions of splitting scores when we know that
the clusters is mono-clonal.
To this end, we simulate reference and alternative read count data of monoclonal
clusters of different sizes (2,3,4 and 5-cell clusters) and add these to the
original dataset, run the tree inference algorithm and compute the splitting
score distributions for all pairs of cells in the same cluster. To ensure that
the simulated data does not confound the tree inference too much, we do this one
cluster at a time.

For each simulated cluster we pick one pair of cells and printed the splitting
score distribution below. With high number of sampled trees, the distributions
of all pairs of cells from the same cluster are very similar, since the model
treats all cells from the same cluster as interchangeable.

Finally, we print the empirical distribution of the the splitting scores for all
clusters of the same size.

The latter is used to specify the cutoff for oligo-clonality: It is defined as
the 95%-percentile of the aggregated distribution of splitting scores.


```{r computing-simulated-clusters, results="hide", dev='png'}
cutoffsSplittingProbs <- data.frame(clusterSize = vector(), Cutoff = vector())
cutoffsBranchingProbabilities <-
  data.frame(clusterSize = vector(), Cutoff = vector())

for (clusterSize in 2:5) {
  try({
    treeNameSimulated <- paste(treeName, clusterSize, sep = "_")


    inputSimulated <- load_data(simulationInputFolder, treeNameSimulated)

    sampleDescriptionSimulated <- inputSimulated$sample_description

    distance <-
      computeClusterSplits(
        inputSimulated$sample_description, inputSimulated$postSampling,
        treeNameSimulated, inputSimulated$nCells, inputSimulated$nMutations,
        inputSimulated$nClusters, inputSimulated$alleleCount,
        inputSimulated$mutatedReadCounts, inputSimulated$totalReadCounts,
        nMutationSamplingEvents = nMutationSamplingEvents,
        nTreeSamplingEvents = nTreeSamplingEvents,
        cellPairSelection = c(
          "orchid", "orchid1", "orchid2",
          "orchid3", "orchid4", "darkorchid",
          "darkorchid1", "darkorchid2", "darkorchid3",
          "darkorchid4", "purple", "purple1",
          "purple2", "purple3", "purple4"
        )
      )


    cutoffsSplittingProbs <-
      rbind(
        cutoffsSplittingProbs,
        data.frame(
          clusterSize = clusterSize,
          Cutoff =
            mean(distance$splittingProbs$Splitting_probability) +
              2 * sd(distance$splittingProbs$Splitting_probability)
        )
      )

    plot(
      ggplot(
        data.frame(x = distance$aggregatedBranchingProbabilities), aes(x = x)
      ) +
        geom_histogram(binwidth = 0.01)
    )

    cutoffsBranchingProbabilities <-
      rbind(
        cutoffsBranchingProbabilities,
        data.frame(
          clusterSize = clusterSize,
          Cutoff =
            quantile(distance$aggregatedBranchingProbabilities,
              probs = 0.95, names = FALSE
            )[1]
        )
      )
  })
}
print(cutoffsBranchingProbabilities)
```



Now we can compute the aggregated splitting score distributions for each
cluster. The distribution's mean is compared to the cutoffs computed above, and
if it is higher than the cutoff, we call the cluster oligo-clonal.

```{r computing-real-clusters}
nTumorClusters <- 0
nOligoclonalClusters2 <- 0
splittingSummary2 <-
  data.frame(Color = vector(), Oligoclonal = vector(), ClusterSize = vector())

for (clusterSize in 2:5) {
  try({
    clusterColor <- input$sample_description %>%
      filter(WBC == 0 & color != "gray93") %>%
      group_by(color) %>%
      filter(n() == clusterSize) %>%
      pull(color) %>%
      unique()

    for (color in clusterColor) {
      distance <-
        computeClusterSplits(input$sample_description, input$postSampling,
          treeName, input$nCells, input$nMutations,
          input$nClusters, input$alleleCount,
          input$mutatedReadCounts, input$totalReadCounts,
          nMutationSamplingEvents = nMutationSamplingEvents,
          nTreeSamplingEvents = nTreeSamplingEvents,
          cellPairSelection = c(color)
        )

      splittingProbs <- mean(distance$splittingProbs$Splitting_probability)
      branchingProbs <- mean(distance$aggregatedBranchingProbabilities)

      nTumorClusters <- nTumorClusters + 1
      oligoclonal <- FALSE

      if (
        branchingProbs >
          cutoffsBranchingProbabilities[
            (cutoffsBranchingProbabilities$clusterSize == clusterSize), 2
          ]
      ) {
        nOligoclonalClusters2 <- nOligoclonalClusters2 + 1
        oligoclonal <- TRUE
      }
      splittingSummary2 <-
        rbind(
          splittingSummary2,
          data.frame(
            Color = color,
            Oligoclonal = oligoclonal, ClusterSize = clusterSize,
            cutoff =
              cutoffsBranchingProbabilities[
                (cutoffsBranchingProbabilities$clusterSize ==
                  clusterSize), 2
              ], mean_splitting_score = branchingProbs
          )
        )
    }
  })
}


numberOfCancerClusters <- input$sample_description %>%
  filter(WBC == 0 & color != "gray93") %>%
  group_by(color) %>%
  filter(n() > 1) %>%
  pull(color) %>%
  unique() %>%
  length()

print(
  sprintf(
    "%d out of %d clusters were found to be oligoclonal in %s, using method 2",
    nOligoclonalClusters2, numberOfCancerClusters, treeName
  )
)
print(splittingSummary2)
library(readr)
write_csv(
  splittingSummary2,
  file.path(inputFolder, paste("splittingSummary_", treeName, ".csv", sep = ""))
)
```
