---
title: "Top separating mutations for CTC cluster cell pairs"
author: "Katharina Jahn, Johannes Gawron"
date: "March 2024"
output:
  html_document:
    keep_md: yes
---

## Index

1. Data
2. Method (explained based on LM2)
3. Results for other cases


## Data 

```{r initialization}
source('../software/Rcode/annotateVariants.R')
sampleName <- 'Br11'
inputFolder <- '../input_folder/'
```

#### Mutation distance matrix
For each cluster (defined by color), we computed a pairwise distance for each mutation pair that indicates how often the two mutations occur in the same private branch of cells from the cluster:

	dist(M1, M2) = 0 (for M1 = M2)
	dist(M1,M2) = 1 - (%samples where M1 and M2 are both in the same private branch of a cell from the cluster) (elsewise)
	
A **private branch** is defined as the path from a leaf to the node just below the LCA of this leaf to another leaf from the same cluster.

This is a generalization of the earlier method to find the top seperating mutations of pairs of leafs. The generalization was necessary to handle the larger clusters that were broken in more than 2 pieces.

```{r}
clusterName <- 'lightcoral'

d <- read.table(file.path(inputFolder, sampleName, paste0(sampleName, '_postSampling_',clusterName,'.txt') ),header=TRUE,sep="\t", stringsAsFactors=F, row.names=1)
mat<-as.matrix(d)
mat[1:4, 1:4]
```

#### Position-wise coverage score
For each position, we computed the percentage of samples that have a coverage of at least 3 at this position. This is meant as a simple score of the data quality of a position that can be used in addition to the separation score to pick mutations for the wet lab experiments. Furthermore, we added simple functional annotations to the variants.

```{r message=FALSE}
coverage<-read.table(file.path(inputFolder, sampleName, paste(sampleName, 'covScore.txt', sep = '_')),header=TRUE,sep="\t", stringsAsFactors=F, row.names=1)
coverage$variantName <- rownames(coverage)
head(coverage)
annotations <- annotate_variants(sampleName, inputFolder)

coverage <- inner_join(coverage, annotations, by = "variantName")

```

## Method
#### Mutation clustering
1. Overview: Raw plot of the distance matrix.
2. Filter distant mutations: Remove all mutations that are not close to any other mutations (minDist>0.5)
3. Dendrogram: Use the distance matrix to cluster the mutations using hierarchical clustering.
4. Cluster remaining mutations: Re-do the hierarchical clustering witht the remaining mutations
5. Define cut point to get about as many groups as there are cluster pieces
6. Rank top separating mutations: Within each group, reduce distance matrix to mutations in the group, rank them by their average distance to other mutations in the group.


###Overview
To get an overview, we plot the full distance matrix:

```{r raw plot}
library(heatmaply)

heatmaply(mat)
```


### Filter out distant mutations

```{r}
mat2 <- mat
diag(mat2) <- 1
min_dist <- apply(mat2, 1, min) # find minimum distance to other mutations
selected_muts <- which(min_dist<0.9) # select those below 0.5 say
mat2 <- mat[selected_muts, selected_muts]
```


This is what the distance matrix looks like now:

```{r}
heatmaply(mat2)
```

```{r}
coverage %>% filter(variantName %in% colnames(mat2))
```


### Dendrogram of the remaining mutations

To cluster mutations, we create a dendrogram based on the pairwise distances:
```{r}
d_mat <- as.dist(mat)
hc <- hclust(d_mat, "average")                   ## hierarchical clustering of mutations based on distance matrix
par(cex=0.6)
plot(hc, main = "Dendrogram based on average pairwise distance", sub = "", xlab = "Separating mutations")
```


No apparent clustering visible.