---
title: "Top separating mutations for CTC cluster cell pairs"
author: "Katharina Jahn, Johannes Gawron"
date: "March 2024"
output:
  html_document:
    keep_md: yes
---

## Index

1. Data
2. Method (explained based on LM2)
3. Results for other cases


## Data 

```{r initialization, message = FALSE}
source("../../workflow/resources/annotateVariants.R")
sample_name <- "Brx50"
input_folder <- "/cluster/work/bewi/members/jgawron/projects/CTC/input_folder"

annotations <- annotate_variants(sample_name, input_folder)
```

#### Mutation distance matrix
For each cluster (defined by color), we computed a pairwise distance for each
mutation pair that indicates how often the two mutations occur in the same
private branch of cells from the cluster:

	dist(M1, M2) = 0 (for M1 = M2)
	dist(M1,M2) = 1 - (%samples where M1 and M2 are both in the same private
	branch of a cell from the cluster) (elsewise)
	
A **private branch** is defined as the path from a leaf to the node just below
the LCA of this leaf to another leaf from the same cluster.

This is a generalization of the earlier method to find the top separating
mutations of pairs of leafs. The generalization was necessary to handle the
larger clusters that were broken in more than 2 pieces.


# lightcoral

```{r}
cluster_name <- "lightcoral"

d <- read.table(
  file.path(
    input_folder, sample_name,
    paste0(sample_name, "_postSampling_", cluster_name, ".txt")
  ),
  header = TRUE,
  sep = "\t", stringsAsFactors = FALSE, row.names = 1
)
mat <- as.matrix(d)
mat[1:4, 1:4]
```

#### Position-wise coverage score
For each position, we computed the percentage of samples that have a coverage of
at least 3 at this position. This is meant as a simple score of the data quality
of a position that can be used in addition to the separation score to pick
mutations for the wet lab experiments. Furthermore, we added simple functional
annotations to the variants.

```{r message=FALSE}
coverage <- read.table(
  file.path(
    input_folder, sample_name, paste(sample_name, "covScore.txt", sep = "_")
  ),
  header = TRUE, sep = "\t", stringsAsFactors = FALSE, row.names = 1
)
coverage$variantName <- rownames(coverage)
head(coverage)

coverage <- inner_join(coverage, annotations, by = "variantName")
```

## Method
#### Mutation clustering
1. Overview: Raw plot of the distance matrix.
2. Filter distant mutations: Remove all mutations that are not close to any
other mutations (minDist>0.5)
3. Dendrogram: Use the distance matrix to cluster the mutations using
hierarchical clustering.
4. Cluster remaining mutations: Re-do the hierarchical clustering with the
remaining mutations
5. Define cut point to get about as many groups as there are cluster pieces
6. Rank top separating mutations: Within each group, reduce distance matrix to
mutations in the group, rank them by their average distance to other mutations
in the group.


###Overview
To get an overview, we plot the full distance matrix:

```{r raw plot}
library(heatmaply)

heatmaply(mat)
```
### Filter out distant mutations

```{r}
mat2 <- mat
diag(mat2) <- 1
min_dist <- apply(mat2, 1, min) # find minimum distance to other mutations
selected_muts <- which(min_dist < 0.5) # select those below 0.5 say
mat3 <- mat[selected_muts, selected_muts]
```

This is what the distance matrix looks like now:

```{r}
heatmaply(mat3)
```


### Dendrogram of the remaining mutations

To cluster mutations, we create a dendrogram based on the pairwise distances:
```{r}
mat <- mat3
d_mat <- as.dist(mat)
hc <- hclust(d_mat, "average") ## hierarchical clustering of mutations based on
# distance matrix
par(cex = 0.6)
plot(
  hc,
  main = "Dendrogram based on average pairwise distance", sub = "",
  xlab = "Separating mutations"
)
```
No apparent clustering structure visible.


# sandybrown 

```{r}
cluster_name <- "sandybrown"

d <- read.table(
  file.path(
    input_folder, sample_name,
    paste0(sample_name, "_postSampling_", cluster_name, ".txt")
  ),
  header = TRUE,
  sep = "\t", stringsAsFactors = FALSE, row.names = 1
)
mat <- as.matrix(d)
mat[1:4, 1:4]
```

#### Position-wise coverage score
For each position, we computed the percentage of samples that have a coverage of
at least 3 at this position. This is meant as a simple score of the data quality
of a position that can be used in addition to the separation score to pick
mutations for the wet lab experiments. Furthermore, we added simple functional
annotations to the variants.

```{r message=FALSE}
coverage <- read.table(
  file.path(
    input_folder, sample_name, paste(sample_name, "covScore.txt", sep = "_")
  ),
  header = TRUE, sep = "\t", stringsAsFactors = FALSE, row.names = 1
)
coverage$variantName <- rownames(coverage)
head(coverage)

coverage <- inner_join(coverage, annotations, by = "variantName")
```

## Method
#### Mutation clustering
1. Overview: Raw plot of the distance matrix.
2. Filter distant mutations: Remove all mutations that are not close to any
other mutations (minDist>0.5)
3. Dendrogram: Use the distance matrix to cluster the mutations using
hierarchical clustering.
4. Cluster remaining mutations: Re-do the hierarchical clustering with the
remaining mutations
5. Define cut point to get about as many groups as there are cluster pieces
6. Rank top separating mutations: Within each group, reduce distance matrix to
mutations in the group, rank them by their average distance to other mutations
in the group.


###Overview
To get an overview, we plot the full distance matrix:

```{r}
library(heatmaply)

heatmaply(mat)
```
### Filter out distant mutations

```{r}
mat2 <- mat
diag(mat2) <- 1
min_dist <- apply(mat2, 1, min) # find minimum distance to other mutations
selected_muts <- which(min_dist < 0.8) # select those below 0.5 say
mat3 <- mat[selected_muts, selected_muts]
```

This is what the distance matrix looks like now:

```{r}
heatmaply(mat3)
```


### Dendrogram of the remaining mutations

To cluster mutations, we create a dendrogram based on the pairwise distances:
```{r}
mat <- mat3
d_mat <- as.dist(mat)
hc <- hclust(d_mat, "average") ## hierarchical clustering of mutations based on
# distance matrix
par(cex = 0.6)
plot(
  hc,
  main = "Dendrogram based on average pairwise distance", sub = "",
  xlab = "Separating mutations"
)
```
No apparent clustering structure visible.
