---
title: "Top separating mutations for CTC cluster cell pairs"
author: "Katharina Jahn, Johannes Gawron"
date: "March 2024"
output:
  html_document:
    keep_md: yes
---

## Index

1. Data
2. Method
3. Results for other cases


## Data 

```{r initialization, message = FALSE}
source("../../workflow/resources/annotateVariants.R")
sample_name <- "Pr9"
inputFolder <- "/cluster/work/bewi/members/jgawron/projects/CTC/input_folder"

annotations <- annotate_variants(sample_name, inputFolder)
```

#### Mutation distance matrix
For each cluster (defined by color), we computed a pairwise distance for each mutation pair that indicates how often the two mutations occur in the same private branch of cells from the cluster:

	dist(M1, M2) = 0 (for M1 = M2)
	dist(M1,M2) = 1 - (%samples where M1 and M2 are both in the same private branch of a cell from the cluster) (elsewise)
	
A **private branch** is defined as the path from a leaf to the node just below the LCA of this leaf to another leaf from the same cluster.

This is a generalization of the earlier method to find the top seperating mutations of pairs of leafs. The generalization was necessary to handle the larger clusters that were broken in more than 2 pieces.


# lightcoral 

```{r}
clusterName <- "lightcoral"

d <- read.table(file.path(inputFolder, sample_name, paste0(sample_name, "_postSampling_", clusterName, ".txt")), header = TRUE, sep = "\t", stringsAsFactors = F, row.names = 1)
mat <- as.matrix(d)
mat[1:4, 1:4]
```

#### Position-wise coverage score
For each position, we computed the percentage of samples that have a coverage of at least 3 at this position. This is meant as a simple score of the data quality of a position that can be used in addition to the separation score to pick mutations for the wet lab experiments. Furthermore, we added simple functional annotations to the variants.

```{r message=FALSE}
coverage <- read.table(file.path(inputFolder, sample_name, paste(sample_name, "covScore.txt", sep = "_")), header = TRUE, sep = "\t", stringsAsFactors = F, row.names = 1)
coverage$variantName <- rownames(coverage)
head(coverage)

coverage <- inner_join(coverage, annotations, by = "variantName")
```

## Method
#### Mutation clustering
1. Overview: Raw plot of the distance matrix.
2. Filter distant mutations: Remove all mutations that are not close to any other mutations (minDist>0.5)
3. Dendrogram: Use the distance matrix to cluster the mutations using hierarchical clustering.
4. Cluster remaining mutations: Re-do the hierarchical clustering witht the remaining mutations
5. Define cut point to get about as many groups as there are cluster pieces
6. Rank top separating mutations: Within each group, reduce distance matrix to mutations in the group, rank them by their average distance to other mutations in the group.


###Overview
To get an overview, we plot the full distance matrix:

```{r}
library(heatmaply)

heatmaply(mat)
```
### Filter out distant mutations

```{r}
mat2 <- mat
diag(mat2) <- 1
min_dist <- apply(mat2, 1, min) # find minimum distance to other mutations
selected_muts <- which(min_dist < 0.9) # select those below 0.5 say
mat3 <- mat[selected_muts, selected_muts]
```

This is what the distance matrix looks like now:

```{r}
heatmaply(mat3)
```


### Dendrogram of the remaining mutations

To cluster mutations, we create a dendrogram based on the pairwise distances:
```{r}
mat <- mat3
d_mat <- as.dist(mat)
hc <- hclust(d_mat, "average") ## hierarchical clustering of mutations based on distance matrix
par(cex = 0.6)
plot(hc, main = "Dendrogram based on average pairwise distance", sub = "", xlab = "Separating mutations")
```
This looks rather complicated and I won't continue here, as choosing clusters would be arbitrary.  




# sandybrown

This is a 3-cell CTC-cluster, so I am expecting up to 3 distinct genotype clusters.

```{r}
clusterName <- "sandybrown"

d <- read.table(file.path(inputFolder, sample_name, paste0(sample_name, "_postSampling_", clusterName, ".txt")), header = TRUE, sep = "\t", stringsAsFactors = F, row.names = 1)
mat <- as.matrix(d)
mat[1:4, 1:4]
```

#### Position-wise coverage score
For each position, we computed the percentage of samples that have a coverage of at least 3 at this position. This is meant as a simple score of the data quality of a position that can be used in addition to the separation score to pick mutations for the wet lab experiments. Furthermore, we added simple functional annotations to the variants.

```{r message=FALSE}
coverage <- read.table(file.path(inputFolder, sample_name, paste(sample_name, "covScore.txt", sep = "_")), header = TRUE, sep = "\t", stringsAsFactors = F, row.names = 1)
coverage$variantName <- rownames(coverage)
head(coverage)

coverage <- inner_join(coverage, annotations, by = "variantName")
```

## Method
#### Mutation clustering
1. Overview: Raw plot of the distance matrix.
2. Filter distant mutations: Remove all mutations that are not close to any other mutations (minDist>0.5)
3. Dendrogram: Use the distance matrix to cluster the mutations using hierarchical clustering.
4. Cluster remaining mutations: Re-do the hierarchical clustering witht the remaining mutations
5. Define cut point to get about as many groups as there are cluster pieces
6. Rank top separating mutations: Within each group, reduce distance matrix to mutations in the group, rank them by their average distance to other mutations in the group.


###Overview
To get an overview, we plot the full distance matrix:

```{r}
library(heatmaply)

heatmaply(mat)
```
### Filter out distant mutations

```{r}
mat2 <- mat
diag(mat2) <- 1
min_dist <- apply(mat2, 1, min) # find minimum distance to other mutations
selected_muts <- which(min_dist < 0.5) # select those below 0.5 say
mat3 <- mat[selected_muts, selected_muts]
```

This is what the distance matrix looks like now:

```{r}
heatmaply(mat3)
```


### Dendrogram of the remaining mutations

To cluster mutations, we create a dendrogram based on the pairwise distances:
```{r}
mat <- mat3
d_mat <- as.dist(mat)
hc <- hclust(d_mat, "average") ## hierarchical clustering of mutations based on distance matrix
par(cex = 0.6)
plot(hc, main = "Dendrogram based on average pairwise distance", sub = "", xlab = "Separating mutations")
```

### Identifying the clusters

This dataset is complicated and clusterings are highly unstable/depend strongly on the filter.
Nonetheless, for the sake of completeness, we provide a clustering here:

We define a cut point to get distinct branches. These should roughly represent the mutations on the private branches in the posterior sampling. The number of clusters can either be the number of distinct branches in the posterior sampling or decided based on the hierarchical clustering.


```{r}
par(cex = 0.6)
plot(hc, main = "Dendrogram based on average pairwise distance", sub = "", xlab = "Separating mutations")
abline(h = 0.65, lwd = 2, lty = 2, col = "green")
```

```{r}
# geneGroups <- cutree(hc, k = NULL, h = 0.6)
geneGroups <- cutree(hc, k = 3)
```



### Rank mutations within cluster
To rank the mutations within one cluster, we reduce the distance matrix to the cluster mutations and rank them by their average distance to the other mutations in the cluster. 

#### First cluster:

```{r}
cluster1 <- names(geneGroups)[geneGroups == 1]
```

Mutations in cluster:
```{r, echo=FALSE}
cluster1
```

Distances within cluster:
```{r}
d1 <- d[cluster1, cluster1]
```

Average distance to other mutations in cluster:
```{r}
colMeans(d1, na.rm = FALSE, dims = 1)
```
Technically we should exclude the value on the main diagonal from the average. But it is zero at all positions which means when only ordering the mutations based on the average, it makes no difference.

**Top separating mutations (first branch):** 
```{r}
(top_names <- names(sort(colMeans(d1, na.rm = FALSE, dims = 1))))
top_df <- as.data.frame(colMeans(d1[top_names]))
colnames(top_df) <- "avgDist"
top_df$variantName <- rownames(top_df)
```

```{r}
top_muts_1 <- inner_join(top_df, coverage, by = "variantName")
# colnames(top_muts_1)[1] <- "mutation"
top_muts_1
```

```{r}
print(sprintf("Number of mutations in cluster %s with moderate functional impact: %d", clusterName, sum(top_muts_1 == "MODERATE")))
print(sprintf("Number of mutations in cluster %s with high functional impact: %d", clusterName, sum(top_muts_1 == "HIGH")))
```




#### Second cluster:

```{r}
cluster1 <- names(geneGroups)[geneGroups == 2]
```

Mutations in cluster:
```{r, echo=FALSE}
cluster1
```

Distances within cluster:
```{r}
d1 <- d[cluster1, cluster1]
```

Average distance to other mutations in cluster:
```{r}
colMeans(d1, na.rm = FALSE, dims = 1)
```
Technically we should exclude the value on the main diagonal from the average. But it is zero at all positions which means when only ordering the mutations based on the average, it makes no difference.

**Top separating mutations (second branch):** 
```{r}
(top_names <- names(sort(colMeans(d1, na.rm = FALSE, dims = 1))))
top_df <- as.data.frame(colMeans(d1[top_names]))
colnames(top_df) <- "avgDist"
top_df$variantName <- rownames(top_df)
```

```{r}
top_muts_2 <- inner_join(top_df, coverage, by = "variantName")
# colnames(top_muts_1)[1] <- "mutation"
top_muts_2
```

```{r}
print(sprintf("Number of mutations in cluster %s with moderate functional impact: %d", clusterName, sum(top_muts_2 == "MODERATE")))
print(sprintf("Number of mutations in cluster %s with high functional impact: %d", clusterName, sum(top_muts_2 == "HIGH")))
```



#### Third cluster:

```{r}
cluster1 <- names(geneGroups)[geneGroups == 3]
```

Mutations in cluster:
```{r, echo=FALSE}
cluster1
```

Distances within cluster:
```{r}
d1 <- d[cluster1, cluster1]
```

Average distance to other mutations in cluster:
```{r}
colMeans(d1, na.rm = FALSE, dims = 1)
```
Technically we should exclude the value on the main diagonal from the average. But it is zero at all positions which means when only ordering the mutations based on the average, it makes no difference.

**Top separating mutations (second branch):** 
```{r}
(top_names <- names(sort(colMeans(d1, na.rm = FALSE, dims = 1))))
top_df <- as.data.frame(colMeans(d1[top_names]))
colnames(top_df) <- "avgDist"
top_df$variantName <- rownames(top_df)
```

```{r}
top_muts_2 <- inner_join(top_df, coverage, by = "variantName")
# colnames(top_muts_1)[1] <- "mutation"
top_muts_2
```

```{r}
print(sprintf("Number of mutations in cluster %s with moderate functional impact: %d", clusterName, sum(top_muts_2 == "MODERATE")))
print(sprintf("Number of mutations in cluster %s with high functional impact: %d", clusterName, sum(top_muts_2 == "HIGH")))
```
